---
title: "Linear Regression Diagnostics and Remidial Measures"
author: "Parash Upreti"
date: "July 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Consequences of Incorrect Functional Form

Creating data
```{r,fig=T}
set.seed(5366)
x=1:100
X=cbind(1,x,x^2)
beta=c(10,-1.15,0.0205)
y=X%*%beta+5*rnorm(100)
mydata=data.frame(y,x)

#Plot
plot(y~x,data=mydata)

#Fitting Quadratic Model
model2=lm(y~x+I(x^2),data=mydata)
summary(model2)

plot(y~x,data=mydata)

x.plot.vector=1:100
y.plot.vector=predict(model2,
                      newdata=data.frame(x=x.plot.vector))
lines(x.plot.vector,
      y.plot.vector,
      col="red")
```

Residuals for Quadratic Model

```{r,fig=T}
betahat2=coef(model2)
X2=model.matrix(model2)

e2=y-X2%*%betahat2
#e2
e2=residuals(model2)
#e2

plot(x,e2)
lines(x,rep(0,length(x)),col="red")
```


Fitting a linear model

```{r,fig=T}
plot(y~x,data=mydata)

model1=lm(y~x,data=mydata)
summary(model1)

abline(model1,col="red")


#Residuals for Linear Model
e1=residuals(model1)

plot(x,e1)
lines(x,rep(0,length(x)),col="red")
```


f-test Function

```{r}
ftest=function(model0,model){
  d0=length(coef(model0))
  d=length(coef(model))
  n=length(residuals(model))
  F=((deviance(model0)-deviance(model))/(d-d0))/
    ((deviance(model))/(n-d))
  return(pf(F,d-d0,n-d,lower.tail=FALSE))
}

ftest(model1,model2)
summary(model2)
```

Plots involving Yhat for Quadratic Model

```{r,fig=T}
X2 = model.matrix(model2)
betahat2=coef(model2)
yhat2=X2%*%betahat2

plot(yhat2,y)
lines(1:100,1:100,col="red")

plot(yhat2,e2)
lines(1:100,rep(0,100),col="red")
```


Plots involving Yhat for Linear Model

```{r, fig=T}
X1 = model.matrix(model1)
betahat1=coef(model1)
yhat1=X1%*%betahat1

plot(yhat1,y)
lines(1:100,1:100,col="red")

plot(yhat1,e1)
lines(1:100,rep(0,100),col="red")
```



### Design Matrix Assumptions

Rank of Design Matrix for Model2

```{r}
library(Matrix)

head(X2)
dim(X2)

rankMatrix(X2)

#Model without Full Rank
z=x
model3=lm(y~x+I(x^2)+z)

X3=model.matrix(model3)

head(X3)
dim(X3)

rankMatrix(X3)

summary(model3)


#Multicollinearity
w=x+rnorm(100)
plot(x,w)
cor(x,w)

model4=lm(y~x+I(x^2)+w)
summary(model4)
```


### NORMALITY OF ERROR TERMS

Normal Data

```{r,fig=T}
set.seed(1234)
e=rnorm(1000)
hist(e)

qqnorm(e)
shapiro.test(e)

#Uniform Data
e=runif(1000)
hist(e)

qqnorm(e)
shapiro.test(e)


#Model with Nonnormal Errors
x=1:100
y=(200+3*x+20*rnorm(100))^3.56
plot(x,y)

model=lm(y~x)
abline(model,col="red")

e=residuals(model)

qqnorm(e)
shapiro.test(e)
```


Achieving Normality with Box-Cox Transformations

```{r, fig=T}
library(MASS)

boxcox.results=boxcox(model)
cbind(boxcox.results$x,boxcox.results$y)

which.max(boxcox.results$y)
lambda=boxcox.results$x[which.max(boxcox.results$y)]


y.tilde=y^lambda


trans.model=lm(y.tilde~x)

plot(x,y.tilde)
abline(trans.model,col="red")

e.tilde=residuals(trans.model)
qqnorm(e.tilde)
shapiro.test(e.tilde)
```

Using Transformed Model to Make Predictions

```{r, fig=T}
new.data=data.frame(x=runif(10,min=0,max=100))
y.tilde.hat=predict(trans.model,newdata=new.data)

plot(new.data$x,y.tilde.hat)
abline(trans.model,col="red")

y.hat=y.tilde.hat^(1/lambda)

plot(x,y)
points(new.data$x,y.hat,col="red",pch=20)
```

### Constancy of Error Variance (Homoscedasticity)

```{r, fig=T}
library("lawstat")


plot(x,abs(e))
levene.test(e,as.factor(x<=median(x)))


plot(x,abs(e.tilde))
levene.test(e.tilde,as.factor(x<=median(x)))
```

A Permutation Test Example

Creating Data

```{r, fig=T}
set.seed(403)
x1=runif(100,min=0,max=100)
x2=runif(100,min=0,max=100)
epsilon=runif(100,min=-50,max=50)

y=200+3*x1+5*x2+epsilon

mydata=data.frame(y,x1,x2)
plot(mydata)

#Fitting a model and examining residuals
model=lm(y~x1+x2)
summary(model)

e=residuals(model)
hist(e)
qqnorm(e)
shapiro.test(e)
```

Model is `y = beta0 + beta1*x1 + beta2*x2`

Permutation Test for H0: beta2=0

```{r}
N=1000
beta2.star.vect = 1:N


for(i in 1:N){
  temp.x2 = sample(x2,
                   size=length(x2),
                   replace=FALSE)
  temp.model=lm(y~x1+temp.x2)
  temp.betahat=coef(temp.model)
  beta2.star.vect[i]=temp.betahat[3]
}


hist(beta2.star.vect)     #All beta2 values from permutation test are stored in this vector


betahat=coef(model)

beta2hat=betahat[3]   
beta2hat                  #Actual value of beta2hat that we observed
```

How do they compare?

```{r}
range(beta2.star.vect)

perm.pvalue=mean(abs(beta2hat)<=abs(beta2.star.vect))
perm.pvalue
```